---
title: "Aggregate KLC Curves"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Kappa results from altered datasets

## Preliminary steps

Load libraries

```{r, include=FALSE}
# Packages that need to be loaded
library(dplyr)
library(ggplot2)
```

Load files

```{r include=FALSE}
# Load files
datasets <- readRDS("../../files/datasets.rds")
method_names = readRDS("../../files/method_names.rds")
noise_level <- readRDS("../../files/noise.rds")
noise_names <- readRDS("../../files/noise_names.rds")
instances_names = readRDS("../../files/instances_names.rds")
#quartiles_names = c("25", "50", "75", "100")

# Load results
deciles_df <- readRDS("../../results/KLC_plot_deciles2.rds")
#deciles_df <- deciles_df %>% select(-accuracy, -kappa, -dataset_order, -method_order)
#saveRDS(deciles_df, file = "../../results/KLC_plot_deciles2.rds")

#quartiles_df <- readRDS("../../results/KLC_plot_quartiles2.rds")
#quartiles_df <- quartiles_df %>% select(-accuracy, -kappa, -dataset_order, -method_order)
#saveRDS(quartiles_df, file = "../../results/KLC_plot_quartiles2.rds")
```

## Obtain the aggregated curves from data

### Mean KLC curves for each level of noise

`df1` contains the mean kappa loss value of all datasets for each level of noise only.

```{r include=FALSE}
df1 <- deciles_df %>%
    group_by(technique, noise) %>%
    summarize(kappa_loss = round(mean(kappa_loss, na.rm = TRUE), 2)) %>%
    ungroup()
```

#### Plot the results

Plot each technique separately

```{r eval=FALSE, include=FALSE}
for(method in method_names){
  # Filter data for the current technique
  filtered_data <- subset(df1, technique == method)
  
  # Create plot
  p1 <- ggplot(filtered_data, aes(x = noise, y = kappa_loss)) +
  geom_point() +
  geom_line(aes(noise)) +
  labs(x = "Noise", y = "Kappa Loss") +
  ggtitle(paste0("Technique ", method)) +
  theme_bw() +
  scale_y_continuous(limits = c(0.0, 1), breaks = seq(0, 1, by = 0.1))

  # Print plot
  print(p1)
}
```

Plot all techniques overlapping

```{r eval=FALSE, include=FALSE}
# Create plot
p1 <- ggplot(df1, aes(x = noise, y = kappa_loss, color = factor(technique))) +
geom_point() +
geom_line(aes(noise)) +
labs(x = "Noise", y = "Kappa Loss", color = "Technique") +
ggtitle("Kappa Loss Curves by technique and noise") +
theme_bw() +
scale_y_continuous(limits = c(0.0, 1), breaks = seq(0, 1, by = 0.1))

# Print plot
print(p1)
```

Plot as a grid

```{r eval=FALSE, include=FALSE}
# Create plot
p1 <- ggplot(df1, aes(x = noise, y = kappa_loss, color = factor(technique))) +
  geom_point() +
  geom_line(aes(noise)) +
  labs(x = "Noise", y = "Kappa Loss") +
  ggtitle("Kappa Loss Curves by technique and noise") +
  theme_bw() +
  scale_y_continuous(limits = c(0.0, 1), breaks = seq(0, 1, by = 0.1)) +
  facet_wrap(~ technique)

# Print plot
print(p1)
ggsave("../../results/plots/KLC_means2.png", p1, width = 40, height = 40, dpi = 600)
```

### Mean KLC curves for each level of noise and quartile of instances

`df2` contains the mean kappa loss value of all datasets for each decile and level of noise.

```{r include=FALSE}
df2 <- deciles_df %>%
    group_by(technique, noise, percentage) %>%
    summarize(kappa_loss = round(mean(kappa_loss, na.rm = TRUE), 2)) %>%
    ungroup()

#df2_q <- quartiles_df %>%
#    group_by(technique, noise, percentage) %>%
#    summarize(kappa_loss = round(mean(kappa_loss, na.rm = TRUE), 2)) %>%
#    ungroup()

saveRDS(df2, file = "../../results/meanKLC_d2.rds")
#saveRDS(df2_q, file = "../../results/meanKLC_q2.rds")
```

#### Plot the results

Plot each technique separately

```{r eval=FALSE, include=FALSE}
# Create plot
for(method in method_names){
  for(instance in instances_names) {
      # Filter data for the current technique and instance percentage
      filtered_data <- subset(df2, technique == method & percentage == instance)
      
      # Create plot
      p2 <- ggplot(filtered_data, aes(x = noise, y = kappa_loss)) +
      geom_point() +
      geom_line(aes(noise)) +
      labs(x = "Noise", y = "Kappa Loss") +
      ggtitle(paste0("Technique ", method, " for altered percentage ", instance)) +
      theme_bw() +
      scale_y_continuous(limits = c(0.0, 0.5), breaks = seq(0, 1, by = 0.1))
  
    # Print plot
    print(p2)
  }
}
```

Plot all techniques overlapping

```{r echo=TRUE}
for(instance in instances_names) {
  # Filter data for the current instance percentage
  filtered_data <- subset(df2, percentage == instance & noise != 5)
  
  # Create plot
  p2 <- ggplot(filtered_data, aes(x = noise, y = kappa_loss, color = factor(technique))) +
  geom_point() +
  geom_line(aes(noise)) +
  labs(x = "Noise", y = "Kappa Loss", color = "Technique") +
  ggtitle(paste0("Kappa Loss Curves by technique, noise and ", instance, " % of instances altered")) +
  theme_bw() +
  scale_y_continuous(limits = c(0.0, 0.5), breaks = seq(0, 1, by = 0.1))
  
  # Print plot
  print(p2)
}

ggsave("../../results/plots/KLC_techniques_q2.png", p2, width = 40, height = 40, dpi = 600)
```

Plot as grid for deciles

```{r echo=TRUE}
# Create plot
p2 <- ggplot(df2, aes(x = percentage, y = kappa_loss, color = factor(noise))) +
  geom_point() +
  geom_line(aes(percentage)) +
  labs(x = "Noise", y = "Kappa Loss") +
  ggtitle("Kappa Loss Curves by technique, noise and percentage of instances altered") +
  theme_bw() +
  scale_y_continuous(limits = c(0.0, 0.5), breaks = seq(0, 1, by = 0.1)) +
  facet_wrap(~ technique)

# Print plot
print(p2)
ggsave("../../results/plots/KLC_means_deciles2.png", p2, width = 40, height = 40, dpi = 600)
```

Plot as grid for quartiles

```{r eval=FALSE, include=FALSE}
# Create plot
p3 <- ggplot(df2_q, aes(x = percentage, y = kappa_loss, color = factor(noise))) +
  geom_point() +
  geom_line(aes(percentage)) +
  labs(x = "Noise", y = "Kappa Loss") +
  ggtitle("Kappa Loss Curves by technique, noise and percentage of instances altered") +
  theme_bw() +
  scale_y_continuous(limits = c(0.0, 0.5), breaks = seq(0, 1, by = 0.1)) +
  facet_wrap(~ technique)

# Print plot
print(p3)
ggsave("../../results/plots/KLC_means_quartiles2.png", p3, width = 40, height = 40, dpi = 600)
```
