---
title: "Clustering of Datasets"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Preliminary steps

Load libraries

```{r, include=FALSE}
# Packages that need to be loaded
library(dplyr)
library(tidyr)
library(ggplot2)
library(GGally) # extensión de ggplot2
library(factoextra) # visualizacion de los clusters
library(NbClust) # determinar el mejor numero de grupos
library(cluster) # medidas de evaluacion como silhouette
library(flexclust) 
library(smacof) 
library(MASS) 
library(ggrepel) # for non-overlapping text labels
```

Load files

```{r include=FALSE}
# Load files
datasets <- readRDS("../files/datasets.rds")
method_names = readRDS("../files/method_names.rds")
noise_level <- readRDS("../files/noise.rds")
noise_names <- readRDS("../files/noise_names.rds")
instances_names = readRDS("../files/instances_names.rds")
quartiles_names = c("25", "50", "75", "100")

# Load results
deciles_df <- readRDS("../results/KLC_plot_deciles.rds")
#deciles_df <- deciles_df %>% select(-accuracy, -kappa, -dataset_order, -method_order)
#saveRDS(deciles_df, file = "../results/KLC_plot_deciles.rds")

quartiles_df <- readRDS("../results/KLC_plot_quartiles.rds")
#quartiles_df <- quartiles_df %>% select(-accuracy, -kappa, -dataset_order, -method_order)
#saveRDS(quartiles_df, file = "../results/KLC_plot_quartiles.rds")
```

------------------------------------------------------------------------

# Clustering

## Obtain the clustering

### Preprocess the data

Transform the data to generate a dataframe to have datasets as rows

```{r echo=TRUE}
# Transform the data
wide_data <- quartiles_df %>%
  # First aggregate across techniques to get average kappa loss per dataset
  group_by(dataset_name, noise, percentage) %>%
  summarize(avg_kappa_loss = mean(kappa_loss, na.rm = TRUE), .groups = 'drop') %>%
  # Then transform to wide format
  unite("noise_percentage", noise, percentage, sep = "_") %>%
  spread(key = noise_percentage, value = avg_kappa_loss)

# View the transformed data
print(wide_data)
```

### Obtain distance matrix

Compute the distance matrix

```{r include=FALSE}
distance_matrix <- dist(wide_data[, -1], method = "euclidean")
```

### Perform Heriarchical Clustering

Colors for clustering

```{r include=FALSE}
# Create a vector of colors based on the number of clusters
colors <- c("#F7AC36","#BF1F5A", "#A1DF91", "#3151CC", "#FF2F20", "#A44FB2")
```

```{r include=FALSE}
# Create a named vector to map clusters to specific colors
cluster_colors <- c(
  "1" = "#F7AC36",
  "2" = "#BF1F5A",
  "3" = "#A1DF91",
  "4" = "#3151CC", 
  "5" = "#FF2F20"
) #F7AC36, #BF1F5A
```

### Silhouette Analysis to Determine Optimal K

```{r echo=TRUE}
# Perform hierarchical clustering
hclusters <- hclust(distance_matrix, method = "ward.D")

# Test different values of k to find the optimal number of clusters
k_values <- c(2, 3, 4)
silhouette_scores <- numeric(length(k_values))
silhouette_plots <- list()

# Calculate silhouette scores for each k
for (i in seq_along(k_values)) {
  k <- k_values[i]
  clusters_temp <- cutree(hclusters, k = k)
  sil_temp <- silhouette(clusters_temp, dist = distance_matrix)
  silhouette_scores[i] <- mean(sil_temp[, 3])
  
  # Store silhouette plot data
  silhouette_plots[[i]] <- sil_temp
  
  cat("k =", k, "- Average silhouette width:", round(silhouette_scores[i], 3), "\n")
}

# Create a dataframe for plotting
silhouette_df <- data.frame(
  k = k_values,
  avg_silhouette = silhouette_scores
)

# Plot silhouette scores
silhouette_comparison_plot <- ggplot(silhouette_df, aes(x = k, y = avg_silhouette)) +
  geom_line(color = "#3151CC", size = 1.2) +
  geom_point(color = "#3151CC", size = 3) +
  scale_x_continuous(breaks = k_values) +
  labs(x = "Number of Clusters (k)", 
       y = "Average Silhouette Width",
       title = "Silhouette Analysis for Optimal Number of Clusters") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

print(silhouette_comparison_plot)

# Save the silhouette comparison plot
png("../results/plots/silhouette_comparison.png", width = 3000, height = 2000, res = 400)
print(silhouette_comparison_plot)
invisible(dev.off())

# Generate individual silhouette plots for each k
for (i in seq_along(k_values)) {
  k <- k_values[i]
  clusters_temp <- cutree(hclusters, k = k)
  
  png(paste0("../results/plots/silhouette_k", k, ".png"), width = 4000, height = 3000, res = 600)
  
  # Create color palette for this k
  colors_k <- cluster_colors[1:k]
  
  plot(silhouette_plots[[i]], col = colors_k, 
       main = paste("Silhouette Plot (k =", k, ")"))
  invisible(dev.off())
}

# Determine the optimal k based on highest average silhouette width
optimal_k_idx <- which.max(silhouette_scores)
optimal_k <- k_values[optimal_k_idx]
cat("\nOptimal number of clusters based on silhouette analysis: k =", optimal_k, "\n")
cat("Average silhouette width for optimal k:", round(silhouette_scores[optimal_k_idx], 3), "\n")

# Use the optimal k for the main analysis (or keep k=4 if preferred)
k <- 4  # You can change this to optimal_k if you want to use the data-driven choice
clusters <- cutree(hclusters, k = k)
```

### Final Clustering Results

```{r echo=TRUE}
# Print cluster assignments
#print(clusters)

# First get unique datasets in the same order as used for clustering
datasets <- wide_data$dataset_name

# Create the mapping dataframe
datasets_clusters <- data.frame(
  dataset = datasets,
  cluster = clusters
)

# Save the dendrogram with colored rectangles by cluster
png("../results/plots/dendogram_d.png", width = 4000, height = 3000, res = 600)
plot(hclusters, hang = -1, labels = wide_data$dataset_name, 
     main = paste("Hierarchical Grouping (k =", k, ")"), 
     xlab = "Observations", sub = NULL)

# Create colored rectangles with consistent colors per cluster
rect.hclust(hclusters, k = k, border = cluster_colors[as.character(1:k)])
invisible(dev.off())

# Generate silhouette plot with consistent colors
png("../results/plots/silhouette_d.png", width = 4000, height = 3000, res = 600)
sil <- silhouette(clusters, dist = distance_matrix)

# Use the same colors for silhouette plot as for dendrogram
plot(sil, col = cluster_colors[as.character(sort(unique(clusters)))], 
     main = paste("Silhouette Plot (k =", k, ")"))
invisible(dev.off())
```

## Plots

```{r include=FALSE}
# Join cluster assignments with original datas
quartiles_df_with_clusters <- quartiles_df %>%
  left_join(datasets_clusters, by = c("dataset_name" = "dataset"))

# Calculate mean kappa loss for each cluster, noise level, and percentage
cluster_means <- quartiles_df_with_clusters %>%
  group_by(cluster, noise, percentage) %>%
  summarize(kappa_loss = round(mean(kappa_loss, na.rm = TRUE), 2), .groups = 'drop')

print(cluster_means)
```

Create plots for individual dataset

```{r echo=TRUE}
# Aggregate the techniques as well
for(instance in quartiles_names) {
  # Filter data for the current instance percentage
  filtered_data <- subset(quartiles_df_with_clusters, percentage == instance)

  # Aggregate by dataset, noise, and cluster (average across techniques)
  aggregated_data <- filtered_data %>%
    group_by(dataset_name, noise, cluster) %>%
    summarize(kappa_loss = mean(kappa_loss, na.rm = TRUE), .groups = "drop")
  
  # Create plot with consistent colors
  p1 <- ggplot(aggregated_data, aes(x = noise, y = kappa_loss, color = factor(cluster))) +
    geom_point() +
    geom_line(aes(group = dataset_name)) +
    # Use consistent colors based on cluster assignment
    scale_color_manual(values = cluster_colors) +
    labs(x = "Noise", y = "Kappa Loss", color = "Cluster") +
    ggtitle(paste0("Kappa Loss Curves by dataset, noise and ", instance, " % of instances altered")) +
    theme_bw() +
    scale_y_continuous(limits = c(0.0, 0.8), breaks = seq(0, 1, by = 0.1))
  
  # Print plot
  print(p1)
}
```

```{r echo=TRUE}
# Create plots for cluster means
for(instance in quartiles_names) {
  # Filter data for the current instance percentage
  filtered_data <- subset(cluster_means, percentage == instance)
  
  # Create plot with consistent colors
  p2 <- ggplot(filtered_data, aes(x = noise, y = kappa_loss, color = factor(cluster))) +
    geom_point() +
    geom_line(aes(group = cluster)) +
    # Use consistent colors based on cluster assignment
    scale_color_manual(values = cluster_colors) +
    labs(x = "Noise", y = "Kappa Loss", color = "Cluster") +
    ggtitle(paste0("Kappa Loss Curves by cluster, noise and ", instance, " % of instances altered")) +
    theme_bw() +
    scale_y_continuous(limits = c(0.0, 0.8), breaks = seq(0, 1, by = 0.1))
  
  # Print plot
  print(p2)
}
```

```{r echo=TRUE}
# Create an empty list to store plots
plot_list <- list()

# Create all plots and store them in the list
for(i in seq_along(quartiles_names)) {
  instance <- quartiles_names[i]
  
  # Filter data for the current instance percentage
  filtered_dataset_data <- subset(quartiles_df_with_clusters, percentage == instance)
  
  # Aggregate by dataset, noise, and cluster
  aggregated_dataset_data <- filtered_dataset_data %>%
    group_by(dataset_name, noise, cluster) %>%
    summarize(kappa_loss = mean(kappa_loss, na.rm = TRUE), .groups = "drop")
  
  # Get cluster means (already aggregated)
  filtered_cluster_data <- subset(cluster_means, percentage == instance)
  
  # Create combined plot with consistent colors
  combined_plot <- ggplot() +
    # Add dataset_name lines with colors based on their cluster
    geom_line(data = aggregated_dataset_data, 
              aes(x = noise, y = kappa_loss, group = dataset_name, color = factor(cluster)),
              linetype = "solid", alpha = 0.5) +
    geom_point(data = aggregated_dataset_data,
               aes(x = noise, y = kappa_loss, group = dataset_name, color = factor(cluster)),
               alpha = 0.5) +
    
    # Add thicker cluster lines to show the averages
    geom_line(data = filtered_cluster_data,
              aes(x = noise, y = kappa_loss, group = cluster, color = factor(cluster)),
              linewidth = 1.5) +
    geom_point(data = filtered_cluster_data,
               aes(x = noise, y = kappa_loss, group = cluster, color = factor(cluster)),
               size = 3) +
    
    # Set the specific color mapping - consistent with other plots
    scale_color_manual(name = "Cluster", values = cluster_colors) +
    
    # Customize the plot
    scale_y_continuous(limits = c(0.0, 1), breaks = seq(0, 1, by = 0.1)) +
    labs(x = "Noise",
         y = "Kappa Loss",
         title = paste0(instance, "% of instances altered")) +
    theme_bw() +
    theme(legend.position = "right")
  
  # Store plot in list
  plot_list[[i]] <- combined_plot
}

# Arrange all plots in a grid using patchwork
if (requireNamespace("patchwork", quietly = TRUE)) {
  # Using patchwork
  library(patchwork)
  combined_grid <- wrap_plots(plot_list, ncol = 1) + 
    plot_annotation(title = "Kappa Loss Curves by Technique and Cluster")
  print(combined_grid)
  
  # Save the grid plot
  png(filename = "../results/plots/cluster_curves_grid2.png", 
      width = 4000, height = 12000, res = 600)
  print(combined_grid)
  dev.off()
} else {
  # Print plots individually if patchwork is not available
  for (p in plot_list) {
    print(p)
  }
}
```
