---
title: "Clustering of Datasets"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Preliminary steps

Load libraries

```{r, include=FALSE}
# Packages that need to be loaded
library(dplyr)
library(tidyr)
library(ggplot2)
library(GGally) # extensión de ggplot2
library(factoextra) # visualizacion de los clusters
library(NbClust) # determinar el mejor numero de grupos
library(cluster) # medidas de evaluacion como silhouette
library(flexclust) 
library(smacof) 
library(MASS) 
```

Load files

```{r include=FALSE}
# Load files
datasets <- readRDS("../files/datasets.rds")
method_names = readRDS("../files/method_names.rds")
noise_level <- readRDS("../files/noise.rds")
noise_names <- readRDS("../files/noise_names.rds")
instances_names = readRDS("../files/instances_names.rds")
quartiles_names = c("25", "50", "75", "100")

# Load results
deciles_df <- readRDS("../results/KLC_plot_deciles.rds")
#deciles_df <- deciles_df %>% select(-accuracy, -kappa, -dataset_order, -method_order)
saveRDS(deciles_df, file = "../results/KLC_plot_deciles.rds")

quartiles_df <- readRDS("../results/KLC_plot_quartiles.rds")
#quartiles_df <- quartiles_df %>% select(-accuracy, -kappa, -dataset_order, -method_order)
saveRDS(quartiles_df, file = "../results/KLC_plot_quartiles.rds")
```

------------------------------------------------------------------------

# Select datasets

### Auxiliary functions to obtain information about the datasets

Create a function to get the number of instances from a dataset

```{r, include=FALSE}
# Function to get the number of instances from a dataset
get_num_instances <- function(dataset) {
  num_instances <- nrow(dataset)
  return(num_instances)
}
```

Create a function to determine the type of each attribute

```{r, include=FALSE}
# Function to determine the type of each attribute
get_attribute_type <- function(dataset) {
  # Exclude the "class" column
  dataset_subset <- dataset[, !names(dataset) %in% "class"]
  
  # Initialize counts
  numerical_count <- 0
  nominal_count <- 0
  
  # Loop through columns
  for (col in names(dataset_subset)) {
    if (is.numeric(dataset_subset[[col]])) {
      numerical_count <- numerical_count + 1
    } else {
      nominal_count <- nominal_count + 1
    }
  }
  
  # Return counts
  return(list(numerical = numerical_count, nominal = nominal_count))
}
```

Create a function to determine if dataset is binary or multiclass

```{r, include=FALSE}
# Function to determine if dataset is binary or multiclass
is_binary <- function(dataset) {
  unique_classes <- unique(dataset$class)
  num_unique_classes <- length(unique_classes)
  
  if (num_unique_classes == 2) {
    return("binary")
  } else if (num_unique_classes > 2) {
    return("multiclass")
  } else {
    stop("Invalid number of unique classes")
  }
}
```

### Information about the datasets

Create a table that shows all datasets and a summary of their data: **`characteristics_df`**

```{r include=FALSE}
# Create a dataframe to store the summary
  characteristics_df <- data.frame(
  dataset_name = character(0),
  instances_n = numeric(0),
  attributes_num = numeric(0),
  attributes_nom = numeric(0),
  dataset_type = character(0),
  stringsAsFactors = FALSE
)

# Iterate through datasets
for (dataset_name in datasets) {
  # Load dataset
  filename = paste0("../datasets/", dataset_name, ".rds")
  dataset <- readRDS(filename)
  # Calculate summary metrics
  num_instances <- get_num_instances(dataset)
  attribute_types <- get_attribute_type(dataset)
  dataset_type <- is_binary(dataset)
  # Add to summary dataframe
    characteristics_df <- bind_rows(characteristics_df, data.frame(
    dataset_name = dataset_name,
    instances_n = num_instances,
    attributes_num = attribute_types$numerical,
    attributes_nom = attribute_types$nominal,
    dataset_type = dataset_type
  ))
}
  
characteristics_df$dataset_type = ifelse(characteristics_df$dataset_type == "binary", 0, 1)

# Convert all data to numeric
characteristics_df$instances_n <- as.numeric(characteristics_df$instances_n)
characteristics_df$attributes_num <- as.numeric(characteristics_df$attributes_num)
characteristics_df$attributes_nom <- as.numeric(characteristics_df$attributes_nom)
characteristics_df$dataset_type <- as.numeric(characteristics_df$dataset_type)

# Combine attribute columns
#characteristics_df$attributes <- characteristics_df$attributes_num + characteristics_df$attributes_nom

# Print the summary dataframe
print(characteristics_df)

saveRDS(characteristics_df, "../files/clustering/characteristics.rds")
```

# Clustering

Colors for clustering

```{r include=FALSE}
# Create a vector of colors based on the number of clusters
colors <- c("#F7AC36","#BF1F5A", "#A1DF91", "#3151CC", "#FF2F20", "#A44FB2")
```

```{r include=FALSE}
# Create a named vector to map clusters to specific colors
cluster_colors <- c(
  "1" = "#F7AC36",
  "2" = "#BF1F5A",
  "3" = "#A1DF91",
  "4" = "#3151CC", 
  "5" = "#FF2F20"
) #F7AC36, #BF1F5A
```

## Clustering on the original data

#### Scale the data

In **`clusters_scaled`** we have the scaled data of the characteristics

```{r include=FALSE}
# Euclidean distances between the rows
clusters_scaled <- scale(characteristics_df[,-1])

# Add dataset names back
clusters_scaled <- cbind(dataset_name = characteristics_df$dataset_name, clusters_scaled)

# Turn into dataframe
clusters_scaled <- as.data.frame(clusters_scaled)

# Print the scaled dataframe
print(clusters_scaled)

saveRDS(clusters_scaled, "../files/clustering/cl_scaled.rds")
```

#### Obtain euclidean distances matrix

Make sure all necessary data is numeric

```{r include=FALSE}
clusters_scaled$instances_n <- as.numeric(clusters_scaled$instances_n)
clusters_scaled$attributes_num <- as.numeric(clusters_scaled$attributes_num)
clusters_scaled$attributes_nom <- as.numeric(clusters_scaled$attributes_nom)
clusters_scaled$dataset_type <- as.numeric(clusters_scaled$dataset_type)

saveRDS(clusters_scaled, "../files/clustering/cl_scaled.rds")
```

In **`dmatrix`** we have the distance matrix for the scaled data

```{r include=FALSE}
set.seed(1)
# Euclidean distances between the rows
dmatrix <- dist(clusters_scaled[,-1])

# Print the distance matrix
print(dmatrix)

saveRDS(dmatrix, "../files/clustering/dmatrix.rds")
```

### Calculate elbow and silhouette to find optimal number of clusters

#### Elbow for **`clusters_scaled`**

```{r echo=TRUE}
# "silhouette", "wss", "gap_stat"
print(fviz_nbclust(clusters_scaled[,-1], kmeans, method = "wss") + geom_vline(xintercept = 4, linetype = 2) + labs(subtitle = "Elbow method") + theme_minimal())
```

#### Get a dataframe with the dataset's names and row ID's

Stored in **`group_list`**

```{r include=FALSE}
# Create a new table with dataset names and rownames
groups_df <- cbind(ID = rownames(clusters_scaled), dataset_name = clusters_scaled$dataset_name)

# Turn into dataframe
groups_df <- as.data.frame(groups_df)

# Make sure all necessary data is numeric
groups_df$ID <- as.numeric(groups_df$ID)
```

#### Silhouette with each dataframe in **`clusters_scaled`**

```{r echo=TRUE}
# Perform hierarchical clustering
hclusters <- hclust(dmatrix, method = "ward.D")

# Plot dendrogram
plot(hclusters, labels = rownames(clusters_scaled), main = "Heirarchal Grouping", xlab = "Observations", sub = NULL, ylim=c(0,1))

# Divide dendogram into groups
k_list <- c(3, 4, 5, 6)
for(i in k_list) {
  groups <- cutree(hclusters, k = i)
  rect.hclust(hclusters, k = i, border = colors[1:i])
  
  # Add dendogram groups to the data
  # Create column name dynamically using paste0
  column_name <- paste0("hclust", i)
  groups_df[[column_name]] <- groups
  
  # Compute silhouette coefficient for hierarchical clustering
  sil <- silhouette(groups, dmatrix)
  
  # Create silhouette plot
  plot(sil, col = colors[1:i], main = paste0("HClust Silhouette with K = ", i))
}
```

> K = 5 seems to be optimal in this case.

### K-Means with the original characteristics

#### K-Means with each dataframe in **`clusters_scaled`**

```{r echo=TRUE}
# Divide dendogram into groups
# k_list <- c(3, 4, 5, 6)
k_list <- c(3, 4, 5, 6)
for(i in k_list) {
  # Set seed for reproducibility
  set.seed(1)
  
  # Run kmeans clustering with dynamic number of centers
  kmeans <- kmeans(clusters_scaled[,-1], centers = i, nstart = 25)
  
  # Extract the centroids
  centroids <- kmeans$centers
  
  # Convert result of kmeans to kcca from flexclust
  kcca <- as.kcca(kmeans, clusters_scaled[,-1])
  
  # Add cluster labels to the groups dataframe with dynamic name
  column_name <- paste0("kmeans_", i)
  groups_df[[column_name]] <- kmeans$cluster
  
  # Visualize the clusters with custom colors
  cluster_plot <- fviz_cluster(kmeans, data = clusters_scaled[,-1], 
                          geom = "point", ellipse.type = "convex", 
                          ggtheme = theme_minimal())
  
  # Apply the custom color palette
  cluster_plot <- cluster_plot + 
                  scale_color_manual(values = colors[1:i]) + 
                  scale_fill_manual(values = colors[1:i]) +
                  coord_fixed(ratio = 1)
  
  # Display the plot
  print(cluster_plot)
  
  # Compute silhouette coefficient for kmeans clustering
  sil_result <- silhouette(kmeans$cluster, dmatrix, clusters_scaled$dataset_name)
  
  # Create silhouette plot
  plot(sil_result, col = colors[1:i], main = paste0("K = ", i, " Silhouette"))
}
```

> K = 5 is optimal.

##### Store the Centroids and the K-Centroids Cluster Analysis

```{r include=FALSE}
# Save the kmeans for k = 4
saveRDS(kmeans, file = "../files/clustering/kmeans.rds")

# Save the centroids
saveRDS(centroids, file = "../files/clustering/centroids.rds")

# Save the KCCA
saveRDS(kcca, file = "../files/clustering/kcca.rds")
```

### Multi-dimensional scaling plot (MDS) with the original characteristics

#### MDS Shepard plot for each dataframe in **`clusters_scaled`**

```{r echo=FALSE}
# Compute Shepard for the distance matrix
fit_mds <- mds(dmatrix)
plot(fit_mds, plot.type = "Shepard")
```

#### MDS Kruskal's Non-metric Multidimensional Scaling for each dataframe in **`clusters_scaled`**

```{r echo=TRUE}
# Nonmetric MDS
# N rows (objects) x p columns (variables)
# each row identified by a unique row name
fit <- isoMDS(dmatrix, k=2) # k is the number of dim

# Plot solution without labels
x <- fit$points[,1]
y <- fit$points[,2]
plot(x, y, xlab="Coordinate 1", ylab="Coordinate 2",
  main="Kruskal's Non-metric MDS", type="n")
points(x, y)
```

#### Show as table of coordinates

```{r include=FALSE}
# Create a table with the results
coord_df <- data.frame(
  ID = clusters_scaled$dataset_name,
  Coordinate1 = x,
  Coordinate2 = y
)

names(coord_df)[names(coord_df) == 'ID'] <- 'dataset_name'
names(coord_df)[names(coord_df) == 'Coordinate1'] <- 'coordinate_1'
names(coord_df)[names(coord_df) == 'Coordinate2'] <- 'coordinate_2'

# Print the results table
print(coord_df)
```

## Plot original data with coordinates clusters

```{r include=FALSE}
# Merge the dataframes
groups_df <- left_join(groups_df, coord_df, by = "dataset_name")
```

### MDS for K = 4 clusters

```{r echo=FALSE}
# Plot solution
x <- groups_df$coordinate_1
y <- groups_df$coordinate_2

# K = 4
# Create the plot
plot(x, y, xlab="Coordinate 1", ylab="Coordinate 2",
     main="Kruskal's Non-metric MDS", type="n")

# Define colors for each cluster
point_colors <- colors[groups_df$kmeans_4]

# Add points colored by cluster
points(x, y, col=point_colors, pch=16)

# Add text labels
#text(x, y, labels=plot_df[,1], cex=0.7, col=point_colors)
```

### MDS for K = 5 clusters

```{r echo=FALSE}
# Plot solution
x <- groups_df$coordinate_1
y <- groups_df$coordinate_2

# K = 5
# Create the plot
plot(x, y, xlab="Coordinate 1", ylab="Coordinate 2",
     main="Kruskal's Non-metric MDS", type="n")

# Define colors for each cluster
point_colors <- colors[groups_df$kmeans_5]

# Add points colored by cluster
points(x, y, col=point_colors, pch=16)

# Add text labels
#text(x, y, labels=plot_df[,1], cex=0.7, col=point_colors)
```

### MDS for K = 6 clusters

```{r echo=FALSE}
# Plot solution
x <- groups_df$coordinate_1
y <- groups_df$coordinate_2

# K = 5
# Create the plot
plot(x, y, xlab="Coordinate 1", ylab="Coordinate 2",
     main="Kruskal's Non-metric MDS", type="n")
 
point_colors <- colors[groups_df$kmeans_5]

# Add points colored by cluster
points(x, y, col=point_colors, pch=16)

# Add text labels
#text(x, y, labels=plot_df[,1], cex=0.7, col=point_colors)
```

#### Store the data for the plots

```{r include=FALSE}
# Save the groups
saveRDS(groups_df, "../files/clustering/groups.rds")
```

## Aggregate plots by dataset and cluster

### Mean KLC curves for each level of noise and quartile of instances

`df3` contains the mean kappa loss value of all techniques for each decile and level of noise.

```{r include=FALSE}
df3 <- deciles_df %>%
    group_by(dataset_name, noise, percentage) %>%
    summarize(kappa_loss = round(mean(kappa_loss, na.rm = TRUE), 2)) %>%
    ungroup()

df3_q <- quartiles_df %>%
    group_by(dataset_name, noise, percentage) %>%
    summarize(kappa_loss = round(mean(kappa_loss, na.rm = TRUE), 2)) %>%
    ungroup()

#saveRDS(df3, file = "../results/meanKLC_d2.rds")
#saveRDS(df3_q, file = "../results/meanKLC_q2.rds")
```

### Join the cluster information with the KLC data

```{r include=FALSE}
# K = 5 is optimal
df3_q_with_clusters <- df3_q %>%
  left_join(groups_df[, c("dataset_name", "kmeans_5")], by = "dataset_name") %>%
  rename(cluster = kmeans_5)

# Get the dataset KLC results with cluster information
cluster_results <- df3_q_with_clusters[c("cluster", "dataset_name", "noise", "percentage", "kappa_loss")]
print(cluster_results)
```

### Calculate mean kappa loss for each dataset cluster, noise level, and percentage

```{r include=FALSE}
cluster_means <- cluster_results %>%
  group_by(cluster, noise, percentage) %>%
  summarize(kappa_loss = round(mean(kappa_loss, na.rm = TRUE), 2)) %>%
  ungroup()

print(cluster_means)

# Save the results
#saveRDS(cluster_means, file = "../results/cluster_means.rds")
```

```{r echo=TRUE}
for(instance in quartiles_names) {
  # Filter data for the current instance percentage
  filtered_data <- subset(df3_q, percentage == instance)
  
  # Join with cluster information
  filtered_data_with_clusters <- filtered_data %>%
    left_join(groups_df[, c("dataset_name", "kmeans_5")], by = "dataset_name") %>%
    rename(cluster = kmeans_5)
  
  # Create plot with colors based on clusters
  p1 <- ggplot(filtered_data_with_clusters, aes(x = noise, y = kappa_loss, 
                                               color = factor(cluster), 
                                               group = dataset_name)) +
    geom_point() +
    geom_line() +
    scale_color_manual(values = cluster_colors, name = "Cluster") +
    labs(x = "Noise", y = "Kappa Loss") +
    ggtitle(paste0("Kappa Loss Curves by dataset (colored by cluster), noise and ", 
                  instance, " % of instances altered")) +
    theme_bw() +
    scale_y_continuous(limits = c(0.0, 0.8), breaks = seq(0, 1, by = 0.1))
  
  # Print plot
  print(p1)
}
```

```{r echo=TRUE}
for(instance in quartiles_names) {
  # Filter data for the current instance percentage
  filtered_data <- subset(cluster_means, percentage == instance)
  
  # Create plot with consistent cluster colors
  p2 <- ggplot(filtered_data, aes(x = noise, y = kappa_loss, color = factor(cluster))) +
    geom_point() +
    geom_line() +
    scale_color_manual(values = cluster_colors, name = "Cluster") +
    labs(x = "Noise", y = "Kappa Loss") +
    ggtitle(paste0("Kappa Loss Curves by cluster, noise and ", 
                  instance, " % of instances altered")) +
    theme_bw() +
    scale_y_continuous(limits = c(0.0, 0.5), breaks = seq(0, 1, by = 0.1))
  
  # Print plot
  print(p2)
}
```

```{r echo=FALSE}
plot_list <- list()

# Create all plots and store them in the list
for(i in seq_along(quartiles_names)) {
  instance <- quartiles_names[i]
  
  # Filter data for both individual datasets and clusters
  filtered_dataset_data <- subset(df3_q_with_clusters, percentage == instance)
  filtered_cluster_data <- subset(cluster_means, percentage == instance)
  
  # Create combined plot
  combined_dataset_plot <- ggplot() +
    # Add individual dataset lines with colors based on their cluster
    geom_line(data = filtered_dataset_data, 
              aes(x = noise, y = kappa_loss, group = dataset_name, color = factor(cluster)),
              linetype = "solid", alpha = 0.5) +
    geom_point(data = filtered_dataset_data,
               aes(x = noise, y = kappa_loss, group = dataset_name, color = factor(cluster)),
               alpha = 0.5) +
    
    # Add thicker cluster lines to show the averages
    geom_line(data = filtered_cluster_data,
              aes(x = noise, y = kappa_loss, group = cluster, color = factor(cluster)),
              linewidth = 1.5) +
    geom_point(data = filtered_cluster_data,
               aes(x = noise, y = kappa_loss, group = cluster, color = factor(cluster)),
               size = 3) +
    
    # Set the specific color mapping
    scale_color_manual(name = "Cluster", values = cluster_colors) +
    
    # Customize the plot
    scale_y_continuous(limits = c(0.0, 0.8), breaks = seq(0, 1, by = 0.1)) +
    labs(x = "Noise",
         y = "Kappa Loss",
         title = paste0(instance, "% of instances altered")) +
    theme_bw() +
    theme(legend.position = "right")
  
  # Store plot in list
  plot_list[[i]] <- combined_dataset_plot
}

# Arrange all plots in a grid using patchwork or gridExtra
if (requireNamespace("patchwork", quietly = TRUE)) {
  # Using patchwork (preferred if available)
  library(patchwork)
  dataset_combined_grid <- wrap_plots(plot_list, ncol = 2) + 
    plot_annotation(title = "Kappa Loss Curves by Dataset and Cluster")
  print(dataset_combined_grid)
} else if (requireNamespace("gridExtra", quietly = TRUE)) {
  # Using gridExtra as an alternative
  library(gridExtra)
  dataset_grid_title <- textGrob("Kappa Loss Curves by Dataset and Cluster", gp = gpar(fontsize = 14))
  dataset_grid_arranged <- grid.arrange(grobs = plot_list, ncol = 2, top = dataset_grid_title)
  print(dataset_grid_arranged)
} else {
  # If neither package is available, print a message and show plots individually
  message("Please install either 'patchwork' or 'gridExtra' package for grid layout.")
  for (p in plot_list) {
    print(p)
  }
}

# Save the grid plot
ggsave("../results/plots/cluster_curves_grid2.png", dataset_combined_grid, width = 40, height = 40, dpi = 600)

```
