---
title: "Kappa Loss Curves for all datasets and methods"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Kappa results from altered datasets

## Preliminary steps

Load libraries

```{r, include=FALSE}
# Packages that need to be loaded
library(dplyr)
library(ggplot2)
```

Load files

```{r, include=FALSE}
# Load files
datasets <- readRDS("../files/datasets.rds")
method_names = readRDS("../files/method_names.rds")
noise_level <- readRDS("../files/noise.rds")
noise_names <- readRDS("../files/noise_names.rds")
instances_names = readRDS("../files/instances_names.rds")
quartiles_names = c("0", "25", "50", "75", "100")

# Load results
mia_df <- readRDS("../results/most_important_attr/mia_df.rds")
noiseMIA_list <- readRDS("../results/noise/noise_list.rds")
instancesCM_list = readRDS("../results/instances/instancesCM_list.rds")
confusion_list <- readRDS("../results/conf_matrices/confusion_matrices.rds")
```

## Information about the datasets

### Auxiliary functions

Function to get the number of instances from a dataset

```{r, include=FALSE}
# Function to get the number of instances from a dataset
get_num_instances <- function(dataset) {
  num_instances <- nrow(dataset)
  return(num_instances)
}
```

Function to determine the type of each attribute

```{r, include=FALSE}
# Function to determine the type of each attribute
get_attribute_type <- function(dataset) {
  # Exclude the "class" column
  dataset_subset <- dataset[, !names(dataset) %in% "class"]
  
  # Initialize counts
  numerical_count <- 0
  nominal_count <- 0
  
  # Loop through columns
  for (col in names(dataset_subset)) {
    if (is.numeric(dataset_subset[[col]])) {
      numerical_count <- numerical_count + 1
    } else {
      nominal_count <- nominal_count + 1
    }
  }
  
  # Return counts
  return(list(Numerical = numerical_count, Nominal = nominal_count))
}
```

Function to determine if dataset is binary or multiclass

```{r, include=FALSE}
# Function to determine if dataset is binary or multiclass
is_binary <- function(dataset) {
  unique_classes <- unique(dataset$class)
  num_unique_classes <- length(unique_classes)
  
  if (num_unique_classes == 2) {
    return("Binary")
  } else if (num_unique_classes > 2) {
    return("Multiclass")
  } else {
    stop("Invalid number of unique classes")
  }
}
```

Create a table that shows all datasets and a summary of their data

```{r echo=FALSE}
# Create a dataframe to store the summary
summary_df <- data.frame(
  Dataset = character(0),
  Num_Instances = numeric(0),
  Numerical_Attributes = numeric(0),
  Nominal_Attributes = numeric(0),
  Dataset_Type = character(0),
  stringsAsFactors = FALSE
)

# Iterate through datasets
for (dataset_name in datasets) {
  # Load dataset
  filename = paste0("../datasets/", dataset_name, ".rds")
  dataset <- readRDS(filename)
  # Calculate summary metrics
  num_instances <- get_num_instances(dataset)
  attribute_types <- get_attribute_type(dataset)
  dataset_type <- is_binary(dataset)
  # Add to summary dataframe
  summary_df <- bind_rows(summary_df, data.frame(
    Dataset = dataset_name,
    Num_Instances = num_instances,
    Numerical_Attributes = attribute_types$Numerical,
    Nominal_Attributes = attribute_types$Nominal,
    Dataset_Type = dataset_type
  ))
}

# Convert all data to numeric
summary_df$Num_Instances <- as.numeric(summary_df$Num_Instances)
summary_df$Numerical_Attributes <- as.numeric(summary_df$Numerical_Attributes)
summary_df$Nominal_Attributes <- as.numeric(summary_df$Nominal_Attributes)

# Combine attribute columns
#summary_df$Attributes <- summary_df$Numerical_Attributes + summary_df$Nominal_Attributes

# Print the summary dataframe
print(summary_df)
```

## Plotting the result

### Turn confusion matrix list into dataframe

Turn confusion list into a dataframe for easier manipulation

```{r include=FALSE}
# Turn list into dataframe
deciles_df = data.frame(matrix(ncol = 6, nrow = 0))
colnames(deciles_df) = c("dataset_name", "technique", "noise", "percentage", "accuracy", "kappa")

for(dataset in datasets) {
  for(method in method_names) {
    for(noise in noise_names) {
      for(instance in instances_names){
        # Get the values for accuracy and kappa 
        a <- confusion_list[[dataset]][[1]][[method]][[noise]][[instance]]$accuracy
        
        k <- confusion_list[[dataset]][[1]][[method]][[noise]][[instance]]$kappa
        
        # Add row to results dataframe
        deciles_df[nrow(deciles_df) + 1,] = c(dataset, method, noise, instance, a, k)
      }
    }
  }
}
```

```{r include=FALSE}
# Turn list into dataframe
quartiles_df = data.frame(matrix(ncol = 6, nrow = 0))
colnames(quartiles_df) = c("dataset_name", "technique", "noise", "percentage", "accuracy", "kappa")

for(dataset in datasets) {
  for(method in method_names) {
    for(noise in noise_names) {
      for(instance in quartiles_names){
        # Get the values for accuracy and kappa 
        a <- confusion_list[[dataset]][[1]][[method]][[noise]][[instance]]$accuracy
        k <- confusion_list[[dataset]][[1]][[method]][[noise]][[instance]]$kappa
        
        # Add row to results dataframe
        quartiles_df[nrow(quartiles_df) + 1,] = c(dataset, method, noise, instance, a, k)
      }
    }
  }
}
```

### Transform values for accuracy, kappa, noise and instances

#### Make sure accuracy and kappa are numbers

```{r, include=FALSE}
deciles_df$accuracy <- as.numeric(deciles_df$accuracy)
deciles_df$kappa <- as.numeric(deciles_df$kappa)
quartiles_df$accuracy <- as.numeric(quartiles_df$accuracy)
quartiles_df$kappa <- as.numeric(quartiles_df$kappa)
```

```{r, include=FALSE}
# Round the values of accuracy and kappa to two decimals
deciles_df$accuracy <- round(deciles_df$accuracy, digits = 2)
deciles_df$kappa <- round(deciles_df$kappa, digits = 2)
quartiles_df$accuracy <- round(quartiles_df$accuracy, digits = 2)
quartiles_df$kappa <- round(quartiles_df$kappa, digits = 2)
```

#### Turn noise and instances into numbers

```{r, include=FALSE}
deciles_df$percentage <- as.numeric(deciles_df$percentage)
deciles_df <- deciles_df %>% mutate(noise = recode(noise, "noise_0" = "0", "noise_5" = "5", "noise_10" = "10", "noise_20" = "20", "noise_30" = "30", "noise_40" = "40", "noise_50" = "50", "noise_60" = "60", "noise_70" = "70", "noise_80" = "80", "noise_90" = "90", "noise_100" = "100"))
deciles_df$noise <- as.numeric(deciles_df$noise)
```

```{r include=FALSE}
quartiles_df$percentage <- as.numeric(quartiles_df$percentage)
quartiles_df <- quartiles_df %>% mutate(noise = recode(noise, "noise_0" = "0", "noise_5" = "5", "noise_10" = "10", "noise_20" = "20", "noise_30" = "30", "noise_40" = "40", "noise_50" = "50", "noise_60" = "60", "noise_70" = "70", "noise_80" = "80", "noise_90" = "90", "noise_100" = "100"))
quartiles_df$noise <- as.numeric(quartiles_df$noise)
```

#### Store the resulting dataframes

```{r, include=FALSE}
saveRDS(deciles_df, file = "../results/results_deciles.rds")
saveRDS(quartiles_df, file = "../results/results_quartiles.rds")
```

## Plotting the Kappa Loss Curves for each dataset and technique in a grid

#### Obtain the Kappa Loss (in deciles)

```{r include=FALSE}
# Get Kappa loss in order to print Kappa Loss Curves
# Calculate 1 - mean_kappa rounded to two decimals
deciles_df$kappa_loss <- round(1 - deciles_df$kappa, 3)
```

```{r echo=FALSE}
# Create a new column to control the order of datasets
deciles_df$dataset_order <- factor(deciles_df$dataset_name, levels = datasets)

# Create a new column to control the order of methods
deciles_df$method_order <- factor(deciles_df$technique, levels = method_names)

# Create plot
p <- ggplot(deciles_df, aes(x = percentage, y = kappa_loss, color = factor(noise))) +
  geom_point() +
  geom_line(aes(group = factor(noise))) +
  labs(x = "Instances", y = "Kappa", color = "Noise") +
  theme_bw() +
  scale_y_continuous(limits = c(0.0, 1), breaks = seq(0, 1, by = 0.1)) +
  facet_grid(method_order ~ dataset_order, scales = "free")

# Print plot
print(p)
ggsave("../results/plots/KLC_plot_deciles.png", p, width = 40, height = 40, dpi = 600)
```

```{r include=FALSE}
saveRDS(deciles_df, file = "../results/KLC_plot_deciles.rds")
```

#### Obtain the Kappa Loss (in quartiles)

```{r include=FALSE}
# Get Kappa loss in order to print Kappa Loss Curves
# Calculate 1 - mean_kappa rounded to two decimals
quartiles_df$kappa_loss <- round(1 - quartiles_df$kappa, 3)
```

```{r echo=FALSE}
# Create a new column to control the order of datasets
quartiles_df$dataset_order <- factor(quartiles_df$dataset_name, levels = datasets)

# Create a new column to control the order of methods
quartiles_df$method_order <- factor(quartiles_df$technique, levels = method_names)

# Create plot
p <- ggplot(quartiles_df, aes(x = percentage, y = kappa_loss, color = factor(noise))) +
  geom_point() +
  geom_line(aes(group = factor(noise))) +
  labs(x = "Instances", y = "Kappa", color = "Noise") +
  theme_bw() +
  scale_y_continuous(limits = c(0.0, 1), breaks = seq(0, 1, by = 0.1)) +
  facet_grid(method_order ~ dataset_order, scales = "free")

# Print plot
print(p)
ggsave("../results/plots/KLC_plot_quartiles.png", p, width = 40, height = 40, dpi = 600)
```

```{r include=FALSE}
saveRDS(quartiles_df, file = "../results/KLC_plot_quartiles.rds")
```
