---
title: "Kappa Loss Curves"
output:
  html_document: default
  pdf_document: default
date: "2024-03-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Kappa results from altered datasets

## Preliminary steps

Load libraries

```{r, include=FALSE}
# Packages that need to be loaded
pacman::p_load(caret, citation, data.table, dplyr, earth, farff, ggpubr, ggplot2, iml, knitr, rpart, tidyverse , tidyr, xtable)

```

Load files

```{r, include=FALSE}
# Load files
datasets <- readRDS("files/datasets.rds")
method_names = readRDS("files/method_names.rds")
noise_level <- readRDS("files/noise.rds")
noise_names <- readRDS("files/noise_names.rds")
instances_names = readRDS("files/instances_names.rds")
quartiles_names = c("0", "25", "50", "75", "100")

# Load results
mia_df <- readRDS("results/mia_df.rds")
noiseMIA_list <- readRDS("results/noise_list.rds")
instancesCM_list = readRDS("results/instancesCM_list.rds")
confusion_list <- readRDS("results/confusion_matrices.rds")

```

## Information about the datasets

### Auxiliary functions

Function to get the number of instances from a dataset

```{r, include=FALSE}
# Function to get the number of instances from a dataset
get_num_instances <- function(dataset) {
  num_instances <- nrow(dataset)
  return(num_instances)
}

```

Function to determine the type of each attribute

```{r, include=FALSE}
# Function to determine the type of each attribute
get_attribute_type <- function(dataset) {
  # Exclude the "class" column
  dataset_subset <- dataset[, !names(dataset) %in% "class"]
  
  # Initialize counts
  numerical_count <- 0
  nominal_count <- 0
  
  # Loop through columns
  for (col in names(dataset_subset)) {
    if (is.numeric(dataset_subset[[col]])) {
      numerical_count <- numerical_count + 1
    } else {
      nominal_count <- nominal_count + 1
    }
  }
  
  # Return counts
  return(list(Numerical = numerical_count, Nominal = nominal_count))
}

```

Function to determine if dataset is binary or multiclass

```{r, include=FALSE}
# Function to determine if dataset is binary or multiclass
is_binary <- function(dataset) {
  unique_classes <- unique(dataset$class)
  num_unique_classes <- length(unique_classes)
  
  if (num_unique_classes == 2) {
    return("Binary")
  } else if (num_unique_classes > 2) {
    return("Multiclass")
  } else {
    stop("Invalid number of unique classes")
  }
}

```

Create a table that shows all datasets and a summary of their data

```{r echo=FALSE}
# Create a dataframe to store the summary
summary_df <- data.frame(
  Dataset = character(0),
  Num_Instances = numeric(0),
  Numerical_Attributes = numeric(0),
  Nominal_Attributes = numeric(0),
  Dataset_Type = character(0),
  stringsAsFactors = FALSE
)

# Iterate through datasets
for (dataset_name in datasets) {
  # Load dataset
  filename = paste0("datasets/", dataset_name, ".rds")
  dataset <- readRDS(filename)
  # Calculate summary metrics
  num_instances <- get_num_instances(dataset)
  attribute_types <- get_attribute_type(dataset)
  dataset_type <- is_binary(dataset)
  # Add to summary dataframe
  summary_df <- bind_rows(summary_df, data.frame(
    Dataset = dataset_name,
    Num_Instances = num_instances,
    Numerical_Attributes = attribute_types$Numerical,
    Nominal_Attributes = attribute_types$Nominal,
    Dataset_Type = dataset_type
  ))
}

# Convert all data to numeric
summary_df$Num_Instances <- as.numeric(summary_df$Num_Instances)
summary_df$Numerical_Attributes <- as.numeric(summary_df$Numerical_Attributes)
summary_df$Nominal_Attributes <- as.numeric(summary_df$Nominal_Attributes)

# Combine attribute columns
#summary_df$Attributes <- summary_df$Numerical_Attributes + summary_df$Nominal_Attributes

# Print the summary dataframe
print(summary_df)

```

#### <!--# Export as LaTex table -->

```{r eval=FALSE, include=FALSE}
# Convert the dataframe to a LaTeX table and print the LaTeX table
print(xtable(summary_df), type = "latex")

```

#### <!--# Most important attribute information -->

#### <!--# Information on most important attribute per dataset and technique -->

```{r eval=FALSE, include=FALSE}
# Create a new dataframe to store the results
mia_classification <- data.frame(
  dataset_name = character(),
  technique = character(),
  mia_type = character(),
  stringsAsFactors = FALSE
)

# Iterate over rows in mia_df
for (i in 1:nrow(mia_df)) {
  dataset_name <- mia_df$dataset_name[i]
  technique <- mia_df$technique[i]
  mia <- mia_df$most_important_attribute[i]
  
  filename = paste0("datasets/", dataset_name, ".rds")
  dataset <- readRDS(filename)
  attribute_type <- class(dataset[[mia]])
  if(class(dataset[[mia]]) == "factor") {
    attribute_type = "nominal"
  }
  
  # Add row to new dataframe
new_row <- data.frame(dataset_name = dataset_name, technique = technique, mia_type = attribute_type)
  mia_classification <- bind_rows(mia_classification, new_row)
}

# Print the new dataframe
print(mia_classification)

```

```{r eval=FALSE, include=FALSE}
# Function to determine dataset type
get_dataset_type <- function(df) {
  # Check if all MIA types are numerical
  if (all(df$mia_type == "numeric")) {
    return("Numerical")
  } 
  # Check if all MIA types are nominal
  else if (all(df$mia_type == "nominal")) {
    return("Nominal")
  } 
  # If neither all numerical nor all nominal, then mixed
  else {
    return("Mixed")
  }
}

# Initialize an empty dataframe to store results
all_results <- data.frame(dataset_name = character(), dataset_type = character(), stringsAsFactors = FALSE)

# Loop through each dataset
for (dataset in datasets) {
  subset_df <- mia_classification %>%
    filter(dataset_name == dataset)
  
  # Group by dataset_name and apply the get_dataset_type function
  result <- subset_df %>%
    group_by(dataset_name) %>%
    summarise(dataset_type = get_dataset_type(.))
  
  # Append the result to the all_results dataframe
  all_results <- bind_rows(all_results, result)
}

# Print the combined result
print(all_results)

```

#### <!--# Re-roder datasets according to classification obtained from most important attributes -->

```{r eval=FALSE, include=FALSE}
#datasets <- c("analcatdata_authorship", "artificial-characters", "banknote", "blood-transfusion-service-center", "breast-w", "cardiotocography", "climate-model-simulation-crashes", "cmc", "diabetes", "iris", "kc1", "liver-disorders", "mfeat-factors", "mfeat-karhunen", "mfeat-zernike", "ozone-level-8hr", "pc4", "phoneme", "qsar-biodeg", "wdbc", "wilt", "badges2", "credit-g", "eucalyptus", "synthetic_control", "tic-tac-toe", "vowel")

datasets <- c("analcatdata_authorship","banknote", "blood-transfusion-service-center", "breast-w", "cardiotocography", "climate-model-simulation-crashes", "cmc", "diabetes", "kc1", "liver-disorders", "mfeat-factors", "mfeat-karhunen", "mfeat-zernike", "ozone-level-8hr", "pc4", "badges2", "credit-g", "eucalyptus")

```

## Plotting the result

### Turn confusion matrix list into dataframe

#### <!--# Troubleshooting in case of error reading confusion_list -->

```{r eval=FALSE, include=FALSE}
dataset <- "analcatdata_authorship"
method <- "svmLinear"
noise <- "noise_50"
instance <- "50"
a <- confusion_list[[dataset]][[1]][[method]][[noise]][[instance]]$accuracy
k <- confusion_list[[dataset]][[1]][[method]][[noise]][[instance]]$kappa
print(a)
print(k)

```

Turn confusion list into a dataframe for easier manipulation

```{r include=FALSE}
# Turn list into dataframe
results_df = data.frame(matrix(ncol = 6, nrow = 0))
colnames(results_df) = c("dataset_name", "technique", "noise", "percentage", "accuracy", "kappa")

for(dataset in datasets) {
  for(method in method_names) {
    for(noise in noise_names) {
      for(instance in instances_names){
        # Get the values for accuracy and kappa 
        a <- confusion_list[[dataset]][[1]][[method]][[noise]][[instance]]$accuracy
        k <- confusion_list[[dataset]][[1]][[method]][[noise]][[instance]]$kappa
        
        # Add row to results dataframe
        results_df[nrow(results_df) + 1,] = c(dataset, method, noise, instance, a, k)
      }
    }
  }
}

```

```{r include=FALSE}
# Turn list into dataframe
quartiles_df = data.frame(matrix(ncol = 6, nrow = 0))
colnames(quartiles_df) = c("dataset_name", "technique", "noise", "percentage", "accuracy", "kappa")

for(dataset in datasets) {
  for(method in method_names) {
    for(noise in noise_names) {
      for(instance in quartiles_names){
        # Get the values for accuracy and kappa 
        a <- confusion_list[[dataset]][[1]][[method]][[noise]][[instance]]$accuracy
        k <- confusion_list[[dataset]][[1]][[method]][[noise]][[instance]]$kappa
        
        # Add row to results dataframe
        quartiles_df[nrow(quartiles_df) + 1,] = c(dataset, method, noise, instance, a, k)
      }
    }
  }
}

```

### Transform values for accuracy, kappa, noise and instances

#### Make sure accuracy and kappa are numbers

```{r, include=FALSE}
results_df$accuracy <- as.numeric(results_df$accuracy)
results_df$kappa <- as.numeric(results_df$kappa)
quartiles_df$accuracy <- as.numeric(quartiles_df$accuracy)
quartiles_df$kappa <- as.numeric(quartiles_df$kappa)

```

```{r, include=FALSE}
# Round the values of accuracy and kappa to two decimals
results_df$accuracy <- round(results_df$accuracy, digits = 2)
results_df$kappa <- round(results_df$kappa, digits = 2)
quartiles_df$accuracy <- round(quartiles_df$accuracy, digits = 2)
quartiles_df$kappa <- round(quartiles_df$kappa, digits = 2)

```

#### Turn noise and instances into numbers

```{r, include=FALSE}
results_df$percentage <- as.numeric(results_df$percentage)
results_df <- results_df %>% mutate(noise = recode(noise, "noise_0" = "0", "noise_5" = "5", "noise_10" = "10", "noise_20" = "20", "noise_30" = "30", "noise_40" = "40", "noise_50" = "50", "noise_60" = "60", "noise_70" = "70", "noise_80" = "80", "noise_90" = "90", "noise_100" = "100"))
results_df$noise <- as.numeric(results_df$noise)

```

```{r, include=FALSE}
quartiles_df$percentage <- as.numeric(quartiles_df$percentage)
quartiles_df <- quartiles_df %>% mutate(noise = recode(noise, "noise_0" = "0", "noise_5" = "5", "noise_10" = "10", "noise_20" = "20", "noise_30" = "30", "noise_40" = "40", "noise_50" = "50", "noise_60" = "60", "noise_70" = "70", "noise_80" = "80", "noise_90" = "90", "noise_100" = "100"))
quartiles_df$noise <- as.numeric(quartiles_df$noise)

```

#### Store the resulting dataframes

```{r}
saveRDS(results_df, file = "results/results_percentiles.rds")
saveRDS(quartiles_df, file = "results/results_quartiles.rds")

```

## Plotting the Kappa Loss Curves for each dataset and technique in a grid

#### Obtain the Kappa Loss (in deciles)

```{r include=FALSE}
# Get Kappa loss in order to print Kappa Loss Curves
# Calculate 1 - mean_kappa rounded to two decimals
results_df$kappa_loss <- round(1 - results_df$kappa, 3)
```

```{r echo=FALSE}
# Create a new column to control the order of datasets
results_df$dataset_order <- factor(results_df$dataset_name, levels = datasets)

# Create a new column to control the order of methods
results_df$method_order <- factor(results_df$technique, levels = method_names)

# Create plot
p <- ggplot(results_df, aes(x = percentage, y = kappa_loss, color = factor(noise))) +
  geom_point() +
  geom_line(aes(group = factor(noise))) +
  labs(x = "Instances", y = "Kappa", color = "Noise") +
  theme_bw() +
  scale_y_continuous(limits = c(0.0, 1), breaks = seq(0, 1, by = 0.1)) +
  facet_grid(method_order ~ dataset_order, scales = "free")

# Print plot
print(p)
ggsave("results/plots/KLC_plot.png", p, width = 40, height = 40, dpi = 1000)

```

```{r include=FALSE}
saveRDS(results_df, file = "results/results_plot_d.rds")
```

#### Obtain the Kappa Loss (in quartiles)

```{r include=FALSE}
# Get Kappa loss in order to print Kappa Loss Curves
# Calculate 1 - mean_kappa rounded to two decimals
quartiles_df$kappa_loss <- round(1 - quartiles_df$kappa, 3)
```

```{r echo=FALSE}
# Create a new column to control the order of datasets
quartiles_df$dataset_order <- factor(quartiles_df$dataset_name, levels = datasets)

# Create a new column to control the order of methods
quartiles_df$method_order <- factor(quartiles_df$technique, levels = method_names)

# Create plot
p <- ggplot(quartiles_df, aes(x = percentage, y = kappa_loss, color = factor(noise))) +
  geom_point() +
  geom_line(aes(group = factor(noise))) +
  labs(x = "Instances", y = "Kappa", color = "Noise") +
  theme_bw() +
  scale_y_continuous(limits = c(0.0, 1), breaks = seq(0, 1, by = 0.1)) +
  facet_grid(method_order ~ dataset_order, scales = "free")

# Print plot
print(p)
ggsave("results/plots/KLC_plot_q.png", p, width = 40, height = 40, dpi = 1000)
```

```{r include=FALSE}
saveRDS(quartiles_df, file = "results/results_plot_q.rds")
```
